Effects of below-ground space limitation on performance of Eucalyptus seedlings:  Does photosynthesis really control growth?

### 2015-07-20 manuscript/manuscript_pve.Rmd
added in discussion

```{r, results='asis', echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```

@@

Interpreting limitations to plant growth requires understanding of the balance between carbon (C) source and sink activity in order to assess C allocation and biomass partitioning. This study used manipulations of soil volume to test how growth is coupled to physiology, allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally sown seedlings (free). Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that soil volume effect would be largest in the smallest containers, resulting in physical constraints to growth independently of photosynthesis (A). Photosynthesis would then become sink-limited, resulting in the build-up of leaf nonstructural carbohydrates eventually leading to photosynthetic down regulation. We observed a negative container effect on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes mass, partitioning to leaves, stems, roots was conserved after 120 days. Photosynthetic capacity was also significantly reduced in containers, and was related to both leaf nitrogen content and starch accumulation. We developed a seedling growth model that utilized leaf A rates to allocate daily C uptake towards mass growth of stems, leaves and roots. We then asked whether the observed reductions in A explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in the down regulation of A, these reductions were not significant enough to explain observed growth responses. Thus, as photosynthesis and growth were not coordinated an excess pool of non-biomass C resulted in seedlings with soil volume restriction. This research highlights the need to further utilize mass balance approaches when evaluating plant C allocation and confirms that A and growth are not always directly related.

@@

Understanding plant growth requires knowledge of the mass balance that must be achieved between C uptake and subsequent allocation to growth, storage, and respiration. As woody plants have highly integrated systems of competing carbohydrate sinks [@kozlowski1992carbohydrate], growth should principally depend on the allocation of photosynthate among different tissues sinks. At long enough time scales A and growth must be correlated in respect to mass balance, however, at shorter temproal scales growth is not necessarily limited by the availability of recent photosynthate. This has led to the current debate over how strongly plant growth is controlled by either source of sink activity. Consequently, plant growth cannot be simply determined by rates of A making it complex to understand and challenging to model [@fourcaud2008plant]. Despite a wealth of studies large uncertainties still remain regarding the coordination of C supply, via A, and growth of woody species.

@@

\
Alternatively, manipulating plant tissue C sinks is often used to investigate the correlation of A and growth. This is because metabolic signaling networks, relaying information on C and N status of different tissues, can down regulate photosynthetic activity [@paul2001sink]. If this sink inhibition of A occurs then a close coordination between declines in A and growth should be expected. Whether this photosynthetic down regulation is evident in woody species has been tested through fruit removal, girdling, and low temperatures. In these studies, down regulation of A was frequently correlated to carbohydrate accumulation resulting from reduced tissue sink strength [@iglesias2002regulation;@hoch2002altitudinal; @urban2007girdling; @haouari2013fruit]. However, reductions in A were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic], and stomatal limitation [@li2005photosynthesis]. These mixed results are not surprising as we still know little about the pathways in which plants achieve balance between assimilation, storage, and growth across temporal scales [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to large annual fruiting sinks, they tell us little about source-sink coordination in ‘normal’ field-grown woody species.

\
If source or sink manipulations do not lead to equivalent changes in growth then understanding the fate of assimilated carbon becomes even more crucial. As programmed plasticity in C allocation naturally occur as plants growth and develop, this must be accounted for when evaluating induced variation in growth [@wright2002interpreting]. Thus, when C allocation is altered by a treatment, it should be shown whether these allocation patterns differ from common sized untreated plants [@reich2002root;@poorter2012biomass]. Only then can we test if functional balance preservation or optimal foraging occurs from changes in C uptake. For example, seedlings in containers may shift allocation to leaves if root restriction occurs or alter fine root morphology to increase nutrient uptake. In woody species, shifts in allometric relationships of leaves, stems and roots can reveal if growth is ontogenetically constrained or actively adjusted when A is affected. This then allows additional pools of carbon, such as root exudation or changes in tissue respiration, to be accounted for and mass balance to be achieved.

\
An alternative experimental approach is to lower belowground C sink strength in tree seedlings by manipulating rooting volume, by varying the container size. The advantage of this approach is it allows a large range of manipulations, can be easily compared to naturally planted seedlings and mimics natural conditions as seedlings compete for space or reach bedrock. Seedlings undergo many physiological and morphological changes in response to rooting volume, including biomass partitioning, A, water relations, nutrient uptake and respiration [@nesmith1998effect, and references therein]. Inadequate rooting volume may decrease carbon sink strength by progressively restricting root growth [@thomas1991root].  Container size studies frequently have photosynthetic down-regulation, likely as a result of sink limitation [@arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. A meta-analysis by Poorter et al. [-@poorter2012pot] concluded that A is the process likely to be the strongest affected by pot size and may best explain the observed effect on biomass seen in the large number of studies where containers are used. This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower C exchange and assimilate export rates [@robbins1988effect]. However, evidence in support for a trade-off between C storage and growth in trees is, to date, inconclusive [@palacio2014does]. Based on these previous studies, using container size as a sink-strength manipulation can be used to empirically test how growth and A are coordinated.

@@

2). Reduced sink strength was expected to lead to accumulation of leaf non-structural carbohydrates, which is known to inhibit photosynthsis. We therefore expected a correlation between carbohydrate accumulation and photosynthetic capacity as a function of soil volume.

3). Last, the growth model was expected to find agreement between simulated and observed seedling mass, through direct correlation of the effects of limiiting soil volume on A.

@@

## Experimental design
This experiment was located at the Hawkesbury Forest Experiment site in Richmond, NSW, Australia. Plots were located in open cover with a site history that consists of a paddock that was converted from native pasture grasses. Top soils at this site, used for the study, are an alluvial formation of low-fertility sandy loam soils (380 and 108 mg kg^-1^ total N and phosphorus respectively) with low organic matter (0.7%) and low water holding capacity. At this site a soil hard layer exists at ~1.0 m with a transition to heavy clay soils. The climate for the region is classified as sub-humid temperate.

@@

\
Plants were watered weekly or when needed, accounting for natural precipitation, to maintain soil moisture at field capacity (13-15%). Drain systems were built into each pot to prevent pooling of water in containers before root expansion, from reduced root uptake, or from large rainfall events. These conditions could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6~mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed . As small containers (5, 10, and 15 l) have a larger irradiation effect a simple bottom plug was used to drain excess water from the gravel compartment.

## Growth and morphology metrics
Seedlings were planted on January 21^st^ 2013 and stem height, diameter at 15~cm and leaf count were measured weekly thereafter. Once the growth rate of individual plants had significantly declined a full biomass harvest was completed (May 21^st^ 2013). Dry mass of leaves, stems, roots and cumulative leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) was measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing cumulative leaf area by total leaf count of only fully expanded leaves. This value was then used to interpolate cumulative leaf area through time with weekly leaf counts. Root mass was collected by passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to 0.5 m depth.  25 g fresh weight subsamples of washed fine roots were analyzed, using WhinoRhizo software (Regent Instruments Inc.), for specific root length (SRL, cm~m^-1^).

## Photosynthetic parameters
Leaf gas exchange measurements were performed bi-weekly at saturating light (A~sat~) and saturating light and [CO~2~] (A~max~) on new fully expanded leaves. Measurements were initiated only after sufficient new leaf growth occurred (March 17^th^, 2013), approximately 6 weeks following planting, and continued until the biomass harvest. Leaf level gas exchange was measured with a standard leaf chamber equipped with blue-red light emitting diodes using a portable gas exchange system (LI-6400, LI-COR, Lincoln, NE, USA). A~sat~ measurements were made at PPFD of 1800 $\mu$mol m^-1^ s^-1^ and [CO~2~] of 400 $\mu$l l^-1^ and A~max~ with [CO~2~] of 1600 $\mu$l l^-1^ and PPFD of 1800 $\mu$mol photons m^-1^ s^-1^. This choice of light level to achieve light saturation is consistent with other studies on *Eucalyptus* species [@kallarackal1997ecophysiological;@pinkard1998photosynthetic;@crous2013photosynthesis;@drake2014capacity]. These measurements were conducted during midday (10:00-14:00 h) with leaf temperature maintained at 25&deg;C. After leaves acclimated to the chamber environment, net CO~2~ assimilation rate and stomatal conductance (g~s~) were logged 5 times for both A~sat~ and A~max~.

\
Photosynthetic CO~2~ response (AC~i~) curves were also developed at 25&deg;C on a random subset of each container size (n=3) after new leaves were first produced and immediately prior to the final harvest (May 23^rd^ 2013). Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25&deg;C at saturating light (above). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical].

\
Leaf dark respiration rates (R~d~) was measured on each seedling during the same dates as AC~i~ curves using detached leaves inside a conifer chamber attached to the Licor 6400 at least 1 hour after sundown.   Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of R~d~ are standardized rates at 25&deg;C using a Q10 value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

## Leaf water potential
Predawn ($\Psi$~pd~) and midday ($\Psi$~l~) leaf water potentials were measured for each seedling using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA) on fully expanded leaves during the same time period as AC~i~ and R~d~. Leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. $\Psi$~pd~ was measured before sunrise and $\Psi$~l~ at midday 13:00-14:30 h. These measurements were used as a measure of static water stress on the seedlings \citep{sellin1999does}, and to ensure that the bulk soil water availability was high enough for plants as they became larger and roots filled the soil volume.

## Leaf, root and soil chemistry
Leaves used in each gas exchange measurements and subsamples of harvested roots were dried to a constant mass and milled for analysis of N content, $\delta$^13^C, and total non-structural carbohydrates (TNC). Pre-planting soil samples (n=6) and subsamples of soil from each container following harvest were sieved to remove organic material, air dried and milled for analysis. Nitrogen concentrations of leaf and soil samples were determined using a Carlo Erba CE1110 elemental analyzer with thermal conductivity and mass spectromic detection (of N~2~ and CO~2~).  The percentage of N in the sample was calculated by comparison with known standards. Leaf $\delta$^13^C was analyzed  with an Delta V Advantage coupled to a Flash HT and Conflo IV isotope ratio mass spectrometer. Leaf samples were flash combusted at 1000&deg;C to convert to to CO~2~, feed to the mass spectrometer and isotopic signatures are reported relative to the VPDP scale. Leaf TNC concentration was analyzed using a total starch assay kit (Megazyme International 303 Ireland Ltd., Wicklow, Ireland) and includes the starch (mg g^-1^) and soluble sugar (mg g^-1^) concentrations. Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. Specific leaf area (SLA, m^2^ kg^-1^), for leaves sampled during gas exchange campaigns, was then calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass.

@@

\
(1)
$$P_i = L\left(\frac{C_{day'i}~\sigma_s}{\epsilon_c}\right) - R$$
where L is total plant leaf area (m^2^), C~day,i~ is the predicted daily carbon assimilation (g d^-1^), $\sigma$~s~ is a self shading parameter, $epsilon$~c~ is a biomass conversion efficiency parameter and R is the mass based total respiration of all tissue components. Total respiration was calculated as

@@

\
C~day~ was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] in the 'plantecophys' package [@Duursma2014] in R [@RDevelopment] with the mean photosynthetic parameters (J~max~, V~cmax~, R~d~ and g~1~) for each treatment and meteorological data from an onsite weather station. J~max~ and Vc~max~ were estimated from AC~i~ curves (explained above), R~d~ was emprically measured and the g~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2012reconciling] with observed g~s~ values. Examples of the photosynthesis model are described in Medlyn et al. [-@medlyn2002temperature] and the approach of the coupled leaf gas exchange model are described in Duursma et al [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15~m intervals, leaf A rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. C~day~ was calculated by converting predicted rates to mass C gain over 15~min timesteps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of C~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of C~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf A with total plant leaf area. This was accomplished by utilizing 61 previously digitized Eucalyptus seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package in R (fix cite YplantQMC) to build a 3d plant structure based on digitized metrics of plant allometry and crown structure. Inputting the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total A, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf.  The ratio of total A with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedling, independently for each treatment. Next, the linear relationship between $\sigma$~s~ and total leaf area was for determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table.3.

@@

## Gas exchange and photosynthetic parameters
A~sat~ and A~max~ were both significantly higher in the largest container volume and the free seedling at the first measurement campaign (both P<0.001). Across all measurement campaigns A~sat~ (Fig. `r asatfig`) and A~max~ (Table 2.) were consistently higher in free seedlings than in containers (boht P<0.001). The interaction between photosynthetic capacity, leaf starch, and leaf N on a mass basis was marginally significant (P=0.0584) but A~max~ was highly correlated to both leaf N content and leaf starch (both P<0.001). Across all measurement campaign A~max~ was higher when foliar N was also higher, usually associated with low levels of leaf starch (Fig. `r achemfig`a). A~max~  was also lower when leaf starch was high as higher leaf N often did not coincide with high leaf starch (Fig. `r achemfig`b)

\
The photosynthetic parameters J~max~ and Vc~max~ were not different within treatments at the beginning and end of gas exchange campaigns, therefore the parameter means per treatment are reported here (Table. 2).  Overall, both J~max~ and Vc~max~ were significantly higher in free seedlings with little variation between soil volume treatments (P=0.0012 & 0.0021, respectively). Leaf dark respiration rates were not significantly different across soil volumes (Table. 2). The g~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2012reconciling] optimal stomatal conductance model, was lowest in the free seedling and was marginally different across soil volume treatments (Table. 2). Predicted values of g~s~, using the g~1~parameter, where highly correlated with observed values (R^2^= .74, P<0.001, data not shown).

@@

## Modelling seedling biomass
Model M~0~, was able to converge on an optimum LMF (21.6 %) which predicted the mean harvest total biomass of free seedlings within 1.2 %. Using this optimized LMF, the total biomass of modeled seedlings with soil volume restriction (M~1~) was on average 23&pm;2.4 g C more than measured seedlings when comparing against predicted total net leaf C gain (Fig. `r massmodelfig`a). Thus, seedling C mass was overestimated by an average of 50&pm;8.7 % in modeled seedlings across the reductions in daily C assimilation per treatment (Fig. `r massmodelfig`b). As a result, the reductions in leaf A were not sufficient enough to explain the reduction in harvested seedling biomass with soil volume restriction. This resulted in an unexplained pool of excess C, generated from A, that was not allocated to biomass.

@@

```{r airvars, fig.width=6, fig.height=8, echo=FALSE}
source("master_scripts/airvars_plotting.R")
```

>>>>

```{r, results="asis", echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```

@@

Interpreting limitations to plant growth requires understanding of the balance between carbon (C) source and sink activity in order to assess C allocation and biomass partitioning. This study used manipulations of soil volume to test how growth is coupled to physiology, allocation, and sink activity in *Eucalyptus tereticornis* seedlings. We grew seedlings in a large range of container sizes and planted containers flush to the soil alongside naturally sown seedlings (free). Reduced soil volume was expected to induce rapid negative effects on growth and physiology compared to free seedlings. It was hypothesized that soil volume effect would be largest in the smallest containers, resulting in physical constraints to growth independently of photosynthesis (A). Photosynthesis would then become sink-limited, resulting in the build-up of leaf nonstructural carbohydrates eventually leading to photosynthetic down regulation. We observed a negative container effect on aboveground growth soon after the experiment started. Although growth was consistently different across soil volumes mass, partitioning to leaves, stems and roots was conserved after 120 days. Photosynthetic capacity was also significantly reduced in containers, and was related to both leaf nitrogen content and starch accumulation. We developed a seedling growth model that utilized leaf A rates to allocate daily C uptake towards mass growth of stems, leaves and roots. We then asked whether the observed reductions in A explained the observed differences in seedling biomass. We found that although belowground sink limitation resulted in the down regulation of A, these reductions were not significant enough to explain observed growth responses. Thus, as photosynthesis and growth were not coordinated an excess pool of non-biomass C resulted in seedlings with soil volume restriction. This research highlights the need to further utilize mass balance approaches when evaluating plant C allocation and confirms that A and growth are not always directly related.

@@

Understanding plant growth requires knowledge of the mass balance that must be achieved between C uptake and subsequent allocation to growth, storage, and respiration. As woody plants have highly integrated systems of competing carbohydrate sinks [@kozlowski1992carbohydrate], growth should principally depend on the allocation of photosynthate among different tissues and organs. At long enough time scales A and growth must be correlated in respect to mass balance, however, at shorter temporal scales growth is not necessarily limited by the availability of recent photosynthate. This has led to the current debate over how strongly plant growth is controlled by either source of sink activity. Consequently, plant growth cannot be simply determined by rates of A making it complex to understand and challenging to model [@fourcaud2008plant]. Despite a wealth of studies large uncertainties still remain regarding the coordination of C supply, via A, and growth of woody species.

@@

\
Alternatively, manipulating plant tissue C sinks is often used to investigate the correlation of A and growth. This is because metabolic signaling networks, relaying information on C and N status of different tissues, can down regulate photosynthetic activity [@paul2001sink]. If this sink inhibition of A occurs then a close coordination between declines in A and growth should be expected. Whether this photosynthetic down regulation is evident in woody species has been tested through fruit removal, girdling, and low temperatures. In these studies, down regulation of A was frequently correlated to carbohydrate accumulation resulting from reduced tissue sink strength [@iglesias2002regulation;@hoch2002altitudinal; @urban2007girdling; @haouari2013fruit]. However, reductions in A were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic], and stomatal limitation [@li2005photosynthesis]. These mixed results are not surprising as we still know little about the pathways in which plants achieve balance between assimilation, storage, and growth across temporal scales [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to large annual fruiting sinks, they tell us little about source-sink coordination in normal field-grown woody species.

\
If source or sink manipulations do not lead to equivalent changes in growth then understanding the fate of assimilated C becomes even more crucial. As programmed plasticity in C allocation naturally occur as plants growth and develop, this must be accounted for when evaluating induced variation in growth [@wright2002interpreting]. Thus, when C allocation is altered by a treatment, it should be shown whether these allocation patterns differ from common sized untreated plants [@reich2002root;@poorter2012biomass]. Only then can we test if functional balance preservation or optimal foraging occurs from changes in C uptake. For example, seedlings in containers may shift allocation to leaves if root restriction occurs or alter fine root morphology to increase nutrient uptake. In woody species, shifts in allometric relationships of leaves, stems and roots can reveal if growth is ontogenetically constrained or actively adjusted when A is affected. This then allows additional pools of C, such as root exudation or changes in tissue respiration, to be accounted for and mass balance to be achieved.

\
An alternative experimental approach is to lower belowground C sink strength in tree seedlings by manipulating rooting volume, by varying the container size. The advantage of this approach is it allows a large range of manipulations, can be easily compared to naturally planted seedlings and may mimic natural conditions as seedlings compete for space or reach bedrock. Seedlings undergo many physiological and morphological changes in response to rooting volume, including biomass partitioning, A, water relations, nutrient uptake and respiration [@nesmith1998effect, and references therein]. Inadequate rooting volume may decrease C sink strength by progressively restricting root growth [@thomas1991root].  Container size studies frequently have photosynthetic down-regulation, likely as a result of sink limitation [@arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. A meta-analysis by Poorter et al. [-@poorter2012pot] concluded that A is the process likely to be the strongest affected by pot size and may best explain the observed effect on biomass seen in the large number of studies where containers are used. This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower C exchange and assimilate export rates [@robbins1988effect]. However, evidence in support for a trade-off between C storage and growth in trees is, to date, inconclusive [@palacio2014does]. Based on these previous studies, using container size as a sink-strength manipulation can be used to empirically test how growth and A are coordinated.

@@

2). Reduced sink strength was expected to lead to accumulation of leaf non-structural carbohydrates, which is known to inhibit A We therefore expected a correlation between carbohydrate accumulation and photosynthetic capacity as a function of soil volume.

3). Last, the growth model was expected to find agreement between simulated and observed seedling mass, through direct correlation of the effects of limiting soil volume on A.

@@

## Experimental design
This experiment was located at the Hawkesbury Forest Experiment site in Richmond, NSW, Australia. Plots were located in open cover with a site history that consists of a paddock that was converted from native pasture grasses. Top soils at this site, used for the study, are an alluvial formation of low-fertility sandy loam soils (380 and 108 mg kg^-1^ total N and phosphorus respectively) with low organic matter (0.7 %) and low water holding capacity. At this site a soil hard layer exists at ~1.0 m with a transition to heavy clay soils. The climate for the region is classified as sub-humid temperate.

@@

\
Plants were watered weekly or when needed, accounting for natural precipitation, to maintain soil moisture at field capacity (13-15 %). Drain systems were built into each pot to prevent pooling of water in containers before root expansion, from reduced root uptake, or from large rainfall events. These conditions could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6 mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed. As small containers (5, 10, and 15 l) have a larger irradiation effect a simple bottom plug was used to drain excess water from the gravel compartment.

## Growth and morphology metrics
Seedlings were planted on January 21^st^ 2013 and stem height, diameter at 15 cm and leaf count were measured weekly thereafter. Once the growth rate of individual plants had significantly declined a full biomass harvest was completed (May 21^st^ 2013). Dry mass of leaves, stems, roots and cumulative leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) was measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing cumulative leaf area by total leaf count of only fully expanded leaves. This value was then used to interpolate cumulative leaf area through time with weekly leaf counts. Root mass was collected by passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to 0.5 m depth.  25 g fresh weight subsamples of washed fine roots were analyzed, using WhinoRhizo software (Regent Instruments Inc.), for specific root length (SRL, m g^-1^).

## Photosynthetic parameters
Leaf gas exchange measurements were performed bi-weekly at saturating light (A~sat~) and saturating light and [CO~2~] (A~max~) on new fully expanded leaves. Measurements were initiated only after sufficient new leaf growth occurred (March 17^th^, 2013), approximately 6 weeks following planting, and continued until the biomass harvest. Leaf level gas exchange was measured with a standard leaf chamber equipped with blue-red light emitting diodes using a portable gas exchange system (LI-6400, LI-COR, Lincoln, NE, USA). A~sat~ measurements were made at PPFD of 1800 $\mu$mol m^-1^ s^-1^ and [CO~2~] of 400 $\mu$l l^-1^ and A~max~ with [CO~2~] of 1600 $\mu$l l^-1^ and PPFD of 1800 $\mu$mol photons m^-1^ s^-1^. This choice of light level to achieve light saturation is consistent with other studies on *Eucalyptus* species [@kallarackal1997ecophysiological;@pinkard1998photosynthetic;@crous2013photosynthesis;@drake2014capacity]. These measurements were conducted during midday (10:00-14:00 h) with leaf temperature maintained at 25 &deg;C. After leaves acclimated to the chamber environment, net CO~2~ assimilation rate and stomatal conductance (g~s~) were logged 5 times for both A~sat~ and A~max~.

\
Photosynthetic CO~2~ response (AC~i~) curves were also developed at 25 &deg;C on a random subset of each container size (n=3) after new leaves were first produced and immediately prior to the final harvest (May 23^rd^ 2013). Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25 &deg;C at saturating light (above). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical].

\
Leaf dark respiration rates (R~d~) was measured on each seedling during the same dates as AC~i~ curves using detached leaves inside a conifer chamber attached to the Licor 6400 at least 1 hour after sundown.   Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of R~d~ are standardized rates at 25 &deg;C using a Q10 value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

## Leaf water potential
Predawn ($\Psi$~pd~) and midday ($\Psi$~l~) leaf water potentials were measured for each seedling using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA) on fully expanded leaves during the same time period as AC~i~ and R~d~. Leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. $\Psi$~pd~ was measured before sunrise and $\Psi$~l~ at midday 13:00-14:30 h. These measurements were used as a measure of static water stress on the seedlings [@sellin1999does] and to ensure that the bulk soil water availability was high enough for plants as they became larger and roots filled the soil volume.

## Leaf, root and soil chemistry
Leaves used in each gas exchange measurements and subsamples of harvested roots were dried to a constant mass and milled for analysis of N content, $\delta$^13^C, and total non-structural carbohydrates (TNC). Pre-planting soil samples (n=6) and subsamples of soil from each container following harvest were sieved to remove organic material, air dried and milled for analysis. Nitrogen concentrations of leaf and soil samples were determined using a Carlo Erba CE1110 elemental analyzer with thermal conductivity and mass spectromic detection (of N~2~ and CO~2~).  The percentage of N in the sample was calculated by comparison with known standards. Leaf $\delta$^13^C was analyzed  with an Delta V Advantage coupled to a Flash HT and Conflo IV isotope ratio mass spectrometer. Leaf samples were flash combusted at 1000&deg;C to convert to CO~2~, feed to the mass spectrometer and isotopic signatures are reported relative to the VPDP scale.

\
Leaf TNC concentration was analyzed using a total starch assay kit (Megazyme International 303 Ireland Ltd., Wicklow, Ireland) and includes the starch (mg g^-1^) and soluble sugar (mg g^-1^) concentrations. Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. Specific leaf area (SLA, m^2^ kg^-1^), for leaves sampled during gas exchange campaigns, was then calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass.

@@

\
(1)
$$P_i = L\left(\frac{C_{day'i}~\sigma_s}{\epsilon_c}\right) - R$$
where L is total plant leaf area (m^2^), C~day,i~ is the predicted daily carbon assimilation (g d^-1^), $\sigma$~s~ is a self shading parameter, $\epsilon$~c~ is a biomass conversion efficiency parameter and R is the mass based total respiration of all tissue components. Total respiration was calculated as

@@

\
C~day~ was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] in the 'plantecophys' package [@Duursma2014] in R [@RDevelopmentCoreTeam2011] with the mean photosynthetic parameters (J~max~, V~cmax~, R~d~ and g~1~) for each treatment and meteorological data from an onsite weather station. J~max~ and Vc~max~ were estimated from AC~i~ curves (explained above), R~d~ was empirically measured and the g~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2012reconciling] with observed g~s~ values. Examples of the photosynthesis model are described in Medlyn et al. [-@medlyn2002temperature] and the approach of the coupled leaf gas exchange model are described in Duursma et al [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, leaf A rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. C~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of C~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of C~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf A with total plant leaf area. This was accomplished by utilizing 61 previously digitized Eucalyptus seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package in R (fix cite YplantQMC) to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Inputting the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total A, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf.  The ratio of total A with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedling, independently for each treatment. Next, the linear relationship between $\sigma$~s~ and total leaf area was for determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table.3.

@@

## Gas exchange and photosynthetic parameters
A~sat~ and A~max~ were both significantly higher in the largest container volume and the free seedling at the first measurement campaign (both P<0.001). Across all measurement campaigns A~sat~ (Fig. `r asatfig`) and A~max~ (Table 2.) were consistently higher in free seedlings than in containers (both P<0.001). The interaction between photosynthetic capacity, leaf starch, and leaf N on a mass basis was marginally significant (P=0.0584) but A~max~ was highly correlated to both leaf N content and leaf starch (both P<0.001). Across all measurement campaign A~max~ was higher when foliar N was also higher, usually associated with low levels of leaf starch (Fig. `r achemfig`a). A~max~  was also lower when leaf starch was high as higher leaf N often did not coincide with high leaf starch (Fig. `r achemfig`b)

\
The photosynthetic parameters J~max~ and Vc~max~ were not different within treatments at the beginning and end of gas exchange campaigns, therefore the parameter means per treatment are reported here (Table. 2).  Overall, both J~max~ and Vc~max~ were significantly higher in free seedlings with little variation between soil volume treatments (P=0.0012 & 0.0021, respectively). Leaf dark respiration rates were not significantly different across soil volumes (Table. 2). The g~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2012reconciling] optimal stomatal conductance model, was lowest in the free seedling and was marginally different across soil volume treatments (Table. 2). Predicted values of g~s~, using the g~1~ parameter, where highly correlated with observed values (R^2^= .74, P<0.001, data not shown).

@@

## Modelling seedling biomass
Model M~0~, was able to converge on an optimum LMF (21.6 %) which predicted the mean harvest total biomass of free seedlings within 1.2 %. Using this optimized LMF, the total biomass of modeled seedlings with soil volume restriction (M~1~) was on average 23&pm;2.4 g C more than measured seedlings when comparing against predicted total net leaf C gain (Fig. `r massmodelfig`a). Thus, seedling C mass was overestimated by an average of 50&pm;8.7 % in modeled seedlings across the reductions in daily C assimilation per treatment (Fig. `r massmodelfig`b). As a result, the reductions in leaf A were not sufficient enough to explain the reduction in harvested seedling biomass with soil volume restriction. This resulted in an unexplained pool of excess C, generated from A, which was not allocated to biomass.

@@

```{r airvars, fig.width=6, fig.height=8, echo=FALSE, dev="png", dpi=300, dev.args=list(type="cairo")}
source("master_scripts/airvars_plotting.R")
```
